---
title: "Math 150 Survival Analysis Project Test Code"
author: "Carrie Saada"
date: "May 2, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r global_options, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=4, fig.width=5, 
                      fig.align = "center")
library(tidyverse)
library(broom)
library(survival)
library(survminer)
library(coxed)
```

#Importing data:

```{r}
AD = read.csv("AIDSdata.csv")
```


#Trying to do power analysis
```{r}
null = sim.survdata(851, 100, x = 1, beta = 0, censor = (851 - 69)/851)
alt = sim.survdata(851, 100, x= 1, beta = -1.454, censor = (851 - 69)/851)
#alt$data$y
cx = coxph(Surv(y, failed) ~ X, data = alt$data)
#coxph(Surv(y, failed) ~ X, data = null$data)
scx = summary(cx)
pval = scx$waldtest[3]
```

X might be the number of covariates? 

```{r, eval = FALSE}
loopVals = c(1:50)
output = rep(NA, length(loopVals))
#output
for(i in loopVals){
  loopSim = sim.survdata(851, 100, x= 1, beta = -1.454, censor = (851 - 69)/851)
  pval = summary(coxph(Surv(y, failed) ~ X, data = loopSim$data))$waldtest[3]
  output[i] = pval < 0.05
  
}
sum(output) / length(output)
```

```{r, eval = FALSE}
betaVals = c(0, -0.2, -0.5, -1, -1.454, -1.5, -2, -3)
powersForBeta = rep(NA, length(betaVals))
outerLoop = c(1:length(betaVals)) #loop through betas
innerLoop = c(1:1000) #replicates
for(i in outerLoop){
  pBools = rep(NA, length(innerLoop))
  for(j in innerLoop){
    loopSim = sim.survdata(851, 100, x= 1, beta = betaVals[i], censor = (851 - 69)/851)
    pval = summary(coxph(Surv(y, failed) ~ X, data = loopSim$data))$waldtest[3]
    pBools[j] = pval < 0.05
  }
  powersForBeta[i] = sum(pBools) / length(pBools)
}

powersForBeta


```

```{r, eval = FALSE}
popVals = c(100, 400, 800, 851, 1200, 1600, 4000)
powersForPop = rep(NA, length(popVals))
outerLoop = c(1:length(popVals)) #loop through betas
innerLoop = c(1:1000) #replicates
for(i in outerLoop){
  pBools = rep(NA, length(innerLoop))
  for(j in innerLoop){
    loopSim = sim.survdata(popVals[i], 100, x= 1, beta = -1.454, censor = (851 - 69)/851)
    pval = summary(coxph(Surv(y, failed) ~ X, data = loopSim$data))$waldtest[3]
    pBools[j] = pval < 0.05
  }
  powersForPop[i] = sum(pBools) / length(pBools)
}

powersForPop


```

```{r}
popx = c(100, 400, 800, 851, 1200, 1600, 4000)
popy = c(0.314, 0.626, 0.715, 0.724, 0.757, 0.789, 0.869)
plot(popx, popy, main = "Power vs Sample Size", xlab = "Sample Size", ylab = "Power")
```

```{r}
betax = c(0, -0.2, -0.5, -1, -1.454, -1.5, -2, -3)
betay = c(0.036, 0.224, 0.681, 0.787, 0.722, 0.698, 0.641, 0.557)
plot(betax, betay, main = "Power vs Effect Size (beta)", xlab = "Effect Size beta", ylab = "Power")
expbetax = c(1.00000000, 0.81873075, 0.60653066, 0.36787944, 0.23363388, 0.22313016, 0.13533528, 0.04978707)
plot(expbetax, betay, main = "Power vs Effect Size, exp(beta)", xlab = "Effect Size exp(beta)", ylab = "Power")
```



#Some exploratory plots:

```{r}
hist(AD$cd4, main = "Histogram of CD4 counts", xlab = "number of cd4 t-cells")
hist(AD$karnof, main = "Histogram of Karnof scores", xlab = "Karnoff score", breaks = 5)
#I'm having a hard time getting the histogram breaks right - karnoff scores are multiples of 10 and it doesn't like that
plot(AD$age, AD$karnof, main = "plot of karnof score vs age")
plot(AD$cd4, AD$time, main ="time vs cd4 count, ignoring censoring")
```


##Working on COX model

Variables: 
Response: time, censor
Not using: time.d, censor.d (because we're using "aids defining event or death", not just death)
Explanatory: tx, txgrp, strat2, sex, raceth, ivdrug, hemophil, karnof, cd4, priorzdv, age


Let's check if we need interaction. Likelihood ratio test is 2ln(Likelihood of full) - 2ln(Likelihood of reduced)

```{r}
full = coxph(Surv(time, censor) ~ (tx+strat2+sex+raceth+ivdrug+hemophil+karnof+cd4+priorzdv+age)^2, data = AD) 
full$loglik[2]
red = coxph(Surv(time, censor) ~ tx+strat2+sex+raceth+ivdrug+hemophil+karnof+cd4+priorzdv+age, data = AD) 
red$loglik[2]
length(red$coefficients)
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2], df = length(full$coefficients) - length(red$coefficients))
full
```

This is close enough to 0.05 that I'm not comfortable eliminating interaction.  

lets eliminiate hemophil entirely

```{r}
full = coxph(Surv(time, censor) ~ (tx+strat2+sex+raceth+ivdrug+hemophil+karnof+cd4+priorzdv+age)^2, data = AD) 
full$loglik[2]
red = coxph(Surv(time, censor) ~ (tx+strat2+sex+raceth+ivdrug+karnof+cd4+priorzdv+age)^2, data = AD) 
red$loglik[2]
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2] , df = length(full$coefficients) - length(red$coefficients))
#red

```


We can eliminate hemophil entirely. Let's try that with strat 2

```{r}
full = coxph(Surv(time, censor) ~ (tx+strat2+sex+raceth+ivdrug+karnof+cd4+priorzdv+age)^2, data = AD) 
full$loglik[2]
red = coxph(Surv(time, censor) ~ (tx+sex+raceth+ivdrug+karnof+cd4+priorzdv+age)^2, data = AD) 
red$loglik[2]
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2] , df = length(full$coefficients) - length(red$coefficients))
red

```

Okay let's eliminate anything with a p-value of 0.1 or larger. 

```{r}
full = coxph(Surv(time, censor) ~ (tx+sex+raceth+ivdrug+karnof+cd4+priorzdv+age)^2, data = AD)  
full$loglik[2]
red = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*(sex + karnof + age) + ivdrug*(karnof + priorzdv) + karnof*age + cd4, data = AD) 
red$loglik[2]
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2] , df = length(full$coefficients) - length(red$coefficients))
red

```

We can do that, so let's eliminate karnof$*$age and karnof$*$ivdrug

```{r}
full = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*(sex + karnof + age) + ivdrug*(karnof + priorzdv) + karnof*age + cd4, data = AD)
full$loglik[2]
red = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*(sex + karnof + age) + ivdrug*priorzdv + cd4, data = AD) 
red$loglik[2]
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2] , df = length(full$coefficients) - length(red$coefficients))
red

```

Let's try raceth$*$age

```{r}
full = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*sex + raceth*karnof + raceth*age + ivdrug*priorzdv + cd4, data = AD) 
full$loglik[2]
red = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*sex + raceth*karnof + ivdrug*priorzdv + cd4, data = AD)
red$loglik[2]
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2] , df = length(full$coefficients) - length(red$coefficients))
red
```

Let's try raceth$*$karnof

```{r}
full = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*sex + raceth*karnof + ivdrug*priorzdv + cd4, data = AD) 
full$loglik[2]
red = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*sex + karnof + ivdrug*priorzdv + cd4, data = AD)
red$loglik[2]
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2] , df = length(full$coefficients) - length(red$coefficients))
red
```

Okay all the interaction terms are now below 0.05 and the only p-values above 0.05 are variables that are also in interaction terms. So, I have a model. Let's check against the full model with all interaction just to be sure. 
 
```{r}
full = coxph(Surv(time, censor) ~ (tx+strat2+sex+raceth+ivdrug+hemophil+karnof+cd4+priorzdv+age)^2, data = AD) 
red = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*sex + ivdrug*priorzdv + karnof + cd4, data = AD)
1 - pchisq(2 * full$loglik[2] - 2 * red$loglik[2] , df = length(full$coefficients) - length(red$coefficients))
```

Alright that works then. 

The model is:

```{r}
model = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*sex + ivdrug*priorzdv + karnof + cd4, data = AD)
model
```

Let's test removing tx from the model to see what that does to the p-value so I can talk about it in my results. 

```{r}
model = coxph(Surv(time, censor) ~ tx*(sex + priorzdv) + raceth*sex + ivdrug*priorzdv + karnof + cd4, data = AD)
modelNoTx = coxph(Surv(time, censor) ~ raceth*sex + ivdrug*priorzdv + karnof + cd4, data = AD)
1 - pchisq(2 * model$loglik[2] - 2 * modelNoTx$loglik[2] , df = length(model$coefficients) - length(modelNoTx$coefficients))
```

P value of 0.000307816 that model is the same without tx (so model is not the same)